---
title: "Delhi Air Pollution"
subtitle: "STAT-670: Final Project"
author: "Yashkumar Burnwal"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.retina = 2
)
```

```{r libraries}
# Load required libraries
library(tidyverse)
library(lubridate)
library(corrplot)
library(viridis)
library(scales)
library(gridExtra)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(reshape2)
```

------------------------------------------------------------------------

# Load and Inspect the Data

## Read the Data

```{r load-data}
# Read the CSV into a dataframe
pollution <- read.csv("Delhi_6Station_Hourly.csv", stringsAsFactors = FALSE)

# Basic dimensions
cat("Rows:", nrow(pollution), "| Columns:", ncol(pollution), "\n")
cat("Stations:", length(unique(pollution$loc)), "\n")
```

## Data Structure

```{r structure}
str(pollution)
```

## Summary Statistics

```{r summary}
summary(pollution)
```

## First Few Rows

```{r head}
head(pollution) %>%
  kable(caption = "First 6 rows of the dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, font_size = 12)
head(pollution)
```

## Variable Classification

The dataset contains the following variable categories:

### Date/Time Components

-   `year`, `month`, `day`, `hour`

### PM2.5 (Primary Response Variable)

-   `PM2.5` – Fine particulate matter (μg/m³)

### Other Pollutants

| Variable | Description               | Unit  |
|----------|---------------------------|-------|
| PM10     | Coarse particulate matter | μg/m³ |
| NO       | Nitric oxide              | ppb   |
| NO2      | Nitrogen dioxide          | ppb   |
| NOx      | Total nitrogen oxides     | ppb   |
| SO2      | Sulfur dioxide            | ppb   |
| CO       | Carbon monoxide           | mg/m³ |
| Ozone    | Ground-level ozone        | ppb   |
| NH3      | Ammonia                   | ppb   |
| Benzene  | Benzene                   | μg/m³ |

### Meteorological Variables

| Variable | Description         | Unit    |
|----------|---------------------|---------|
| AT       | Ambient Temperature | °C      |
| BP       | Barometric Pressure | hPa     |
| SR       | Solar Radiation     | W/m²    |
| RH       | Relative Humidity   | \%      |
| WS       | Wind Speed          | m/s     |
| WD       | Wind Direction      | degrees |

### Location

-   `loc` – Station name

> **PM2.5 confirmed as the primary target variable**

# Construct Proper Datetime and Clean Data

## Create DateTime Features

We combine the year–month–day–hour columns into a proper POSIXct timestamp and derive useful time-based features.

```{r datetime-features}
pollution <- pollution %>%
  mutate(
    datetime = make_datetime(year, month, day, hour),
    date = as.Date(datetime),
    hour_of_day = hour,                                # 0-23
    dow = wday(datetime, label = TRUE, abbr = FALSE),  # Weekday factor
    month_name = month(datetime, label = TRUE),        # Month factor
    is_weekend = dow %in% c("Saturday", "Sunday")
  )

# Verify datetime range
cat("Date range:", as.character(min(pollution$date)), "to", 
    as.character(max(pollution$date)), "\n")
```

## Missing Values Analysis

```{r missing-values}
# Count missing values per variable
missing_counts <- pollution %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  filter(Missing_Count > 0) %>%
  arrange(desc(Missing_Count))

if(nrow(missing_counts) > 0) {
  missing_counts %>%
    kable(caption = "Variables with Missing Values") %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
} else {
  cat("No missing values found in the dataset.\n")
}

# Rows with missing PM2.5
n_missing_pm25 <- sum(is.na(pollution$PM2.5))
cat("\nRows with missing PM2.5:", n_missing_pm25, "\n")
```

## Outlier Check

```{r outlier-check}
outlier_summary <- pollution %>%
  summarise(
    PM25_negative = sum(PM2.5 < 0, na.rm = TRUE),
    PM25_extreme_high = sum(PM2.5 > 999, na.rm = TRUE),
    AT_below_minus10 = sum(AT < -10, na.rm = TRUE),
    AT_above_55 = sum(AT > 55, na.rm = TRUE),
    RH_out_of_range = sum(RH < 0 | RH > 100, na.rm = TRUE),
    WS_negative = sum(WS < 0, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Check", values_to = "Count")

outlier_summary %>%
  kable(caption = "Outlier Check Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

# Explore PM2.5 Over Time

## Basic Descriptive Statistics

```{r pm25-stats}
pm25_stats <- pollution %>%
  summarise(
    Mean = mean(PM2.5, na.rm = TRUE),
    Median = median(PM2.5, na.rm = TRUE),
    SD = sd(PM2.5, na.rm = TRUE),
    Min = min(PM2.5, na.rm = TRUE),
    Q25 = quantile(PM2.5, 0.25, na.rm = TRUE),
    Q75 = quantile(PM2.5, 0.75, na.rm = TRUE),
    Q90 = quantile(PM2.5, 0.90, na.rm = TRUE),
    Q99 = quantile(PM2.5, 0.99, na.rm = TRUE),
    Max = max(PM2.5, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value") %>%
  mutate(Value = round(Value, 2))

pm25_stats %>%
  kable(caption = "PM2.5 Descriptive Statistics (μg/m³)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Distribution of PM2.5

```{r}
ggplot(pollution, aes(x = PM2.5)) +
  geom_histogram(
    binwidth = 15,
    fill = "steelblue",
    alpha = 0.7,
    color = "white"
  ) +
  geom_vline(
    xintercept = median(pollution$PM2.5, na.rm = TRUE),
    color = "orange",
    linetype = "dashed",
    linewidth = 1
  ) +
  scale_x_continuous(limits = c(0, 500)) +
  labs(
    title = "Frequency vs PM2.5 Range (Binwidth = 15 μg/m³)",
    x = "PM2.5 (μg/m³)",
    y = "Frequency",
    subtitle = "Orange dashed line = median"
  ) +
  theme_minimal(base_size = 12)


```

```{r histogram, fig.cap="Distribution of PM2.5 concentrations with density overlay"}
ggplot(pollution, aes(x = PM2.5)) +
  geom_histogram(aes(y = after_stat(density)), bins = 80, fill = "steelblue", 
                 alpha = 0.7, color = "white") +
  geom_density(color = "darkred", linewidth = 1) +
  geom_vline(xintercept = median(pollution$PM2.5, na.rm = TRUE), 
             color = "orange", linetype = "dashed", linewidth = 1) +
  scale_x_continuous(limits = c(0, 500)) +
  labs(title = "Distribution of PM2.5 Concentrations",
       subtitle = "Orange dashed line = median | Red curve = density",
       x = "PM2.5 (μg/m³)", y = "Density") +
  theme_minimal(base_size = 12)
```

The distribution is **right-skewed**, typical of air pollution data. The long right tail represents pollution episodes, particularly during winter months.

## Full Time Series

```{r timeseries, fig.height=5, fig.cap="Daily mean PM2.5 over the study period"}
# Aggregate to daily means for cleaner visualization
daily_pm25 <- pollution %>%
  group_by(date) %>%
  summarise(
    daily_PM25 = mean(PM2.5, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(daily_pm25, aes(x = date, y = daily_PM25)) +
  geom_line(color = "steelblue", alpha = 0.8) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, color = "darkred", 
              linewidth = 1.2) +
  geom_hline(yintercept = 35, color = "green", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = as.Date("2018-07-01"), y = 45, 
           label = "WHO 24-hr guideline (35 μg/m³)", color = "darkgreen", size = 3) +
  labs(title = "Daily Mean PM2.5 Over Time (June 2018 - October 2019)",
       subtitle = "Red line = LOESS smoother",
       x = "Date", y = "Daily Mean PM2.5 (μg/m³)") +
  theme_minimal(base_size = 12) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b %Y")
```

**Key observations:**

-   Strong **seasonal pattern** with winter peaks (Nov-Jan) and monsoon lows (Jul-Sep)
-   Daily values frequently exceed WHO guideline (35 μg/m³)
-   Winter 2018-19 shows extreme episodes exceeding 300 μg/m³

## Hourly (Diurnal) Cycle

```{r hourly-pattern, fig.cap="Mean PM2.5 by hour of day"}
hourly_pattern <- pollution %>%
  group_by(hour_of_day) %>%
  summarise(
    mean_PM25 = mean(PM2.5, na.rm = TRUE),
    se_PM25 = sd(PM2.5, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(hourly_pattern, aes(x = hour_of_day, y = mean_PM25)) +
  geom_ribbon(aes(ymin = mean_PM25 - 1.96*se_PM25, 
                  ymax = mean_PM25 + 1.96*se_PM25), 
              fill = "steelblue", alpha = 0.3) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "darkblue", size = 2.5) +
  scale_x_continuous(breaks = 0:23) +
  labs(title = "Diurnal Pattern of PM2.5",
       subtitle = "Shaded area = 95% confidence interval",
       x = "Hour of Day (0-23)", y = "Mean PM2.5 (μg/m³)") +
  theme_minimal(base_size = 12)
```

```{r hourly-interpretation}
peak_hour <- hourly_pattern$hour_of_day[which.max(hourly_pattern$mean_PM25)]
low_hour <- hourly_pattern$hour_of_day[which.min(hourly_pattern$mean_PM25)]
cat("Peak PM2.5 hour:", peak_hour, ":00\n")
cat("Lowest PM2.5 hour:", low_hour, ":00\n")
```

**Interpretation:**

-   **Peak hours:** Morning (8-10 AM) and late evening (9-11 PM) - coinciding with rush hour traffic and shallow boundary layer
-   **Minimum:** Afternoon (2-4 PM) - when convective mixing is strongest and pollutants disperse vertically

## Day-of-Week Pattern

```{r dow-pattern, fig.cap="PM2.5 distribution by day of week"}
ggplot(pollution, aes(x = dow, y = PM2.5, fill = dow)) +
  geom_boxplot(outlier.alpha = 0.1, outlier.size = 0.5) +
  scale_y_continuous(limits = c(0, 400)) +
  scale_fill_viridis_d() +
  labs(title = "PM2.5 by Day of Week",
       x = "Day of Week", y = "PM2.5 (μg/m³)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r weekday-weekend}
weekend_compare <- pollution %>%
  group_by(is_weekend) %>%
  summarise(
    Mean_PM25 = mean(PM2.5, na.rm = TRUE),
    Median_PM25 = median(PM2.5, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Period = ifelse(is_weekend, "Weekend", "Weekday"))

weekend_compare %>%
  select(Period, Mean_PM25, Median_PM25) %>%
  kable(caption = "Weekday vs Weekend PM2.5", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

The weekday-weekend difference is **modest** in Delhi compared to Western cities, likely because:

-   Delhi has less rigid work schedules
-   Traffic and commercial activity persist on weekends
-   Dominant pollution sources (dust, burning) are not work-schedule dependent

## Monthly/Seasonal Pattern

```{r monthly-boxplot, fig.height=6, fig.cap="Monthly distribution of PM2.5"}
pollution <- pollution %>%
  mutate(year_month = floor_date(date, "month"))

ggplot(pollution, aes(x = factor(year_month), y = PM2.5)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.alpha = 0.1) +
  scale_y_continuous(limits = c(0, 500)) +
  labs(title = "Monthly Distribution of PM2.5 (June 2018 - October 2019)",
       x = "Month", y = "PM2.5 (μg/m³)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Seasonal Interpretation:**

| Season | Months | PM2.5 Level | Key Drivers |
|------------------|------------------|------------------|------------------|
| **Winter** | Nov–Jan | Very High (200-400+) | Temperature inversions, crop burning, heating emissions, low wind |
| **Pre-Monsoon** | Mar–May | Moderate-High | Dust storms, high temperatures |
| **Monsoon** | Jun–Sep | Low (40-80) | Rain washout, humid air, clean maritime air masses |
| **Post-Monsoon** | Oct–Nov | Rising | Diwali fireworks, crop residue burning begins |

# Explore Relationships with Other Variables

## Correlation Matrix

```{r}
# Select numeric variables
numeric_vars <- pollution %>%
  select(PM2.5, PM10, AT, BP, SR, RH, WS, NO, NO2, SO2, Ozone, CO, Benzene, NH3, NOx) %>%
  na.omit()

# Correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")

# Convert to long format for ggplot2
cor_long <- melt(cor_matrix)

# Pretty heatmap
ggplot(cor_long, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  
  # Gradient colors
  scale_fill_gradient2(
    low = "#8b0000",   # dark red
    mid = "white",
    high = "#003f5c",  # navy blue
    midpoint = 0,
    limit = c(-1, 1),
    name = "Correlation"
  ) +
  
  geom_text(aes(label = sprintf("%.2f", value)), color = "black", size = 3) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    panel.grid = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  ) +
  labs(
    title = "Correlation Heatmap of Pollutants and Meteorology",
    x = "",
    y = ""
  )

```

## Correlations with PM2.5

```{r pm25-correlations}
pm25_correlations <- data.frame(
  Variable = rownames(cor_matrix),
  Correlation = cor_matrix[, "PM2.5"]
) %>%
  filter(Variable != "PM2.5") %>%
  arrange(desc(abs(Correlation))) %>%
  mutate(Correlation = round(Correlation, 3))

# REMOVE rownames to avoid duplicate column in kable
rownames(pm25_correlations) <- NULL

pm25_correlations %>%
  kable(caption = "Correlations with PM2.5 (sorted by absolute value)", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(1:5, bold = TRUE, background = "#f0f8ff")

```

**Top predictors** (highlighted) show strong associations with PM2.5:

-   **PM10** (strongest) - PM2.5 is a subset of PM10; same emission sources
-   **CO, NO2, NOx** - Common combustion sources (vehicles, burning)
-   **AT (negative)** - Cold weather traps pollutants; inversions

## Scatterplots for Top Predictors

```{r scatterplots, fig.height=10, fig.width=12, fig.cap="Scatterplots of PM2.5 vs top predictors"}
# Sample data for cleaner plots
plot_sample <- pollution %>% sample_n(5000)

# PM2.5 vs PM10
p1 <- ggplot(plot_sample, aes(x = PM10, y = PM2.5)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  geom_smooth(method = "loess", color = "darkgreen", se = FALSE, linetype = "dashed") +
  scale_x_continuous(limits = c(0, 600)) +
  scale_y_continuous(limits = c(0, 500)) +
  labs(title = "PM2.5 vs PM10", subtitle = "Red: Linear | Green: LOESS",
       x = "PM10 (μg/m³)", y = "PM2.5 (μg/m³)") +
  theme_minimal()

# PM2.5 vs CO
p2 <- ggplot(plot_sample, aes(x = CO, y = PM2.5)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  geom_smooth(method = "loess", color = "darkgreen", se = FALSE, linetype = "dashed") +
  scale_x_continuous(limits = c(0, 5)) +
  scale_y_continuous(limits = c(0, 500)) +
  labs(title = "PM2.5 vs CO", x = "CO (mg/m³)", y = "PM2.5 (μg/m³)") +
  theme_minimal()

# PM2.5 vs AT (temperature)
p3 <- ggplot(plot_sample, aes(x = AT, y = PM2.5)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", color = "darkred", se = TRUE) +
  scale_y_continuous(limits = c(0, 500)) +
  labs(title = "PM2.5 vs Temperature", subtitle = "Strong nonlinearity at low temps",
       x = "Temperature (°C)", y = "PM2.5 (μg/m³)") +
  theme_minimal()

# PM2.5 vs RH
p4 <- ggplot(plot_sample, aes(x = RH, y = PM2.5)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", color = "darkred", se = TRUE) +
  scale_y_continuous(limits = c(0, 500)) +
  labs(title = "PM2.5 vs Relative Humidity",
       x = "Relative Humidity (%)", y = "PM2.5 (μg/m³)") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

**Observations:**

-   **PM2.5 vs PM10:** Strong positive linear relationship (expected co-emission)
-   **PM2.5 vs CO:** Positive relationship (common combustion sources)
-   **PM2.5 vs AT:** **Nonlinear** - high PM2.5 concentrations cluster at low temperatures (winter inversions)
-   **PM2.5 vs RH:** Complex - moderate humidity can enhance particle growth; very high humidity during monsoon coincides with washout

## Trivariate Visualizations

### PM2.5 vs NO2 by Hour of Day

```{r trivariate-hour, fig.height=7, fig.cap="PM2.5 vs NO2 colored by hour of day"}
ggplot(pollution %>% sample_n(8000), 
       aes(x = NO2, y = PM2.5, color = factor(hour_of_day))) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_viridis_d(option = "plasma") +
  scale_y_continuous(limits = c(0, 400)) +
  scale_x_continuous(limits = c(0, 150)) +
  labs(title = "PM2.5 vs NO2 by Hour of Day",
       subtitle = "Yellow = daytime hours | Purple = nighttime hours",
       x = "NO2 (ppb)", y = "PM2.5 (μg/m³)", color = "Hour") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

The hour-colored plot reveals that **morning hours (8-10 AM)** show elevated NO2-PM2.5 relationships - consistent with traffic rush hour emissions.

### PM2.5 vs Relative Humidity by Season

```{r trivariate-season, fig.height=8, fig.cap="PM2.5 vs RH by season, colored by temperature"}
# Create season variable
pollution <- pollution %>%
  mutate(season = case_when(
    month %in% c(12, 1, 2) ~ "Winter",
    month %in% c(3, 4, 5) ~ "Pre-Monsoon",
    month %in% c(6, 7, 8, 9) ~ "Monsoon",
    month %in% c(10, 11) ~ "Post-Monsoon"
  ),
  season = factor(season, levels = c("Monsoon", "Post-Monsoon", "Winter", "Pre-Monsoon")))

ggplot(pollution %>% sample_n(10000), 
       aes(x = RH, y = PM2.5, color = AT)) +
  geom_point(alpha = 0.4, size = 1) +
  facet_wrap(~season, ncol = 2) +
  scale_color_viridis_c(option = "magma") +
  scale_y_continuous(limits = c(0, 500)) +
  labs(title = "PM2.5 vs Relative Humidity by Season",
       subtitle = "Color gradient shows temperature",
       x = "Relative Humidity (%)", y = "PM2.5 (μg/m³)", color = "Temp (°C)") +
  theme_minimal(base_size = 12)
```

**Insights from trivariate plot:**

-   **Winter:** Extreme PM2.5 across all humidity levels; cold temperatures (purple) dominate
-   **Monsoon:** Universally low PM2.5 due to rain washout effect
-   **Post-Monsoon:** Transition period with rising pollution as crop burning begins
-   The **temperature gradient** clearly separates seasons and pollution regimes

------------------------------------------------------------------------

# Training/Testing Split and Modeling

## Define Train/Test Split

```{r train-test-split}
train_start <- as.Date("2018-06-01")
train_end <- as.Date("2019-06-01")
test_start <- as.Date("2019-06-02")
test_end <- as.Date("2019-10-01")

pollution_train <- pollution %>% filter(date >= train_start & date <= train_end)
pollution_test <- pollution %>% filter(date >= test_start & date <= test_end)

cat("Training period:", as.character(train_start), "to", as.character(train_end), "\n")
cat("Training observations:", format(nrow(pollution_train), big.mark = ","), "\n\n")
cat("Testing period:", as.character(test_start), "to", as.character(test_end), "\n")
cat("Testing observations:", format(nrow(pollution_test), big.mark = ","), "\n")
```

## Aggregate to Daily Dataset

For interpretable modeling, we aggregate hourly data to daily means.

```{r daily-aggregation}
daily_train <- pollution_train %>%
  group_by(date) %>%
  summarise(
    daily_PM25 = mean(PM2.5, na.rm = TRUE),
    daily_PM10 = mean(PM10, na.rm = TRUE),
    daily_NO2 = mean(NO2, na.rm = TRUE),
    daily_NOx = mean(NOx, na.rm = TRUE),
    daily_CO = mean(CO, na.rm = TRUE),
    daily_SO2 = mean(SO2, na.rm = TRUE),
    daily_Ozone = mean(Ozone, na.rm = TRUE),
    daily_AT = mean(AT, na.rm = TRUE),
    daily_RH = mean(RH, na.rm = TRUE),
    daily_WS = mean(WS, na.rm = TRUE),
    daily_SR = mean(SR, na.rm = TRUE),
    daily_BP = mean(BP, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    dow = wday(date, label = TRUE),
    month = month(date)
  )

daily_test <- pollution_test %>%
  group_by(date) %>%
  summarise(
    daily_PM25 = mean(PM2.5, na.rm = TRUE),
    daily_PM10 = mean(PM10, na.rm = TRUE),
    daily_NO2 = mean(NO2, na.rm = TRUE),
    daily_NOx = mean(NOx, na.rm = TRUE),
    daily_CO = mean(CO, na.rm = TRUE),
    daily_SO2 = mean(SO2, na.rm = TRUE),
    daily_Ozone = mean(Ozone, na.rm = TRUE),
    daily_AT = mean(AT, na.rm = TRUE),
    daily_RH = mean(RH, na.rm = TRUE),
    daily_WS = mean(WS, na.rm = TRUE),
    daily_SR = mean(SR, na.rm = TRUE),
    daily_BP = mean(BP, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    dow = wday(date, label = TRUE),
    month = month(date)
  )

cat("Daily training observations:", nrow(daily_train), "days\n")
cat("Daily testing observations:", nrow(daily_test), "days\n")
```

## Multiple Linear Regression

We fit a multiple linear regression using the strongest predictors identified from correlation analysis.

```{r linear-model}
model_lm <- lm(daily_PM25 ~ daily_PM10 + daily_NO2 + daily_NOx + daily_CO + 
                            daily_AT + daily_RH + daily_WS + daily_Ozone,
               data = daily_train)

summary(model_lm)
```

### Model Interpretation

```{r model-interpretation}
# Extract key metrics
r_squared <- summary(model_lm)$r.squared
adj_r_squared <- summary(model_lm)$adj.r.squared
coef_df <- broom::tidy(model_lm) %>%
  mutate(
    significant = p.value < 0.05,
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    statistic = round(statistic, 2),
    p.value = format.pval(p.value, digits = 3)
  )

coef_df %>%
  kable(caption = "Linear Regression Coefficients") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(which(coef_df$significant), bold = TRUE)
```

```{r r-squared}
cat("R-squared:", round(r_squared, 4), "\n")
cat("Adjusted R-squared:", round(adj_r_squared, 4), "\n")
cat("\nThe model explains", round(r_squared * 100, 1), "% of variance in daily PM2.5.\n")
```

**Coefficient interpretations:**

| Variable | Sign | Physical Interpretation |
|------------------------|------------------------|------------------------|
| PM10 | \+ | Expected; PM2.5 is a subset of PM10 |
| AT | − | Higher temperature = better mixing = lower PM2.5 |
| WS | − | Higher wind speed = better dispersion |
| Ozone | − | Photochemical regime; high ozone often indicates cleaner conditions |
| RH | \+ | Hygroscopic growth increases particle mass |

### Diagnostic Plots

```{r diagnostics, fig.height=8, fig.width=10, fig.cap="Regression diagnostic plots"}
par(mfrow = c(2, 2))
plot(model_lm)
par(mfrow = c(1, 1))
```

**Diagnostic assessment:**

-   **Residuals vs Fitted:** Some heteroscedasticity at higher values (expected for pollution data)
-   **Q-Q Plot:** Reasonable normality with slight deviation in tails
-   **Scale-Location:** Confirms mild heteroscedasticity
-   **Residuals vs Leverage:** A few influential points but no extreme outliers

## Test Set Evaluation

```{r test-evaluation}
# Predictions from base model
predictions_lm <- predict(model_lm, newdata = daily_test)

# Handle any NAs
valid_idx <- !is.na(predictions_lm) & !is.na(daily_test$daily_PM25)
predictions_lm_valid <- predictions_lm[valid_idx]
actual_test <- daily_test$daily_PM25[valid_idx]

# Calculate metrics
rmse_lm <- sqrt(mean((predictions_lm_valid - actual_test)^2))
mae_lm <- mean(abs(predictions_lm_valid - actual_test))
mape_lm <- mean(abs((predictions_lm_valid - actual_test) / actual_test)) * 100

# Training metrics for comparison
predictions_train <- predict(model_lm, newdata = daily_train)
rmse_train <- sqrt(mean((predictions_train - daily_train$daily_PM25)^2, na.rm = TRUE))
mae_train <- mean(abs(predictions_train - daily_train$daily_PM25), na.rm = TRUE)
```

```{r metrics-table}
metrics_df <- data.frame(
  Metric = c("RMSE (μg/m³)", "MAE (μg/m³)", "MAPE (%)"),
  Training = c(round(rmse_train, 2), round(mae_train, 2), NA),
  Test = c(round(rmse_lm, 2), round(mae_lm, 2), round(mape_lm, 2))
)

metrics_df %>%
  kable(caption = "Model Performance Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Predicted vs Actual Plot

```{r}
test_results <- data.frame(
  actual = actual_test,
  predicted = predictions_lm_valid
)

ggplot(test_results, aes(x = actual, y = predicted)) +
  
  # Blue dots = predictions vs actual
  geom_point(aes(color = "Observed Points"), alpha = 0.6, size = 2.5) +
  
  # Red dashed line = perfect prediction line
  geom_abline(aes(color = "Perfect Prediction Line"), 
              intercept = 0, slope = 1,
              linetype = "dashed", linewidth = 1) +
  
  # Green line = fitted regression line
  geom_smooth(aes(color = "Fitted Trend Line"), 
              method = "lm", se = TRUE, linewidth = 0.8) +
  
  scale_color_manual(
    name = "Legend",
    values = c(
      "Observed Points" = "steelblue",
      "Perfect Prediction Line" = "red",
      "Fitted Trend Line" = "darkgreen"
    )
  ) +
  
  labs(
    title = "Predicted vs Actual PM2.5 (Test Set: July–October 2019)",
    subtitle = paste0("RMSE = ", round(rmse_lm, 2),
                      " μg/m³ | MAE = ", round(mae_lm, 2), " μg/m³"),
    x = "Actual Daily PM2.5 (μg/m³)",
    y = "Predicted Daily PM2.5 (μg/m³)"
  ) +
  
  coord_fixed(ratio = 1) +
  theme_minimal(base_size = 12)

```

### Prediction Time Series

```{r prediction-timeseries, fig.cap="Time series of actual vs predicted PM2.5"}
daily_test$predicted <- NA
daily_test$predicted[valid_idx] <- predictions_lm_valid

ggplot(daily_test, aes(x = date)) +
  geom_line(aes(y = daily_PM25, color = "Actual"), linewidth = 1) +
  geom_line(aes(y = predicted, color = "Predicted"), linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "steelblue", "Predicted" = "darkred")) +
  labs(title = "Daily PM2.5: Actual vs Predicted (Test Period: June-Oct 2019)",
       x = "Date", y = "Daily Mean PM2.5 (μg/m³)", color = "") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```
